---
title: "Analysis of Wildlife Strikes in Commercial Aviation"
subtitle: "Practicum 1 CS5200"
author: "Rajiv Ragavan"
email: "ragavan.r@northeastern.edu"
date: "Fall 2024"
output: html_document
---

## Loading of Packages
```{r setup, eval = T, echo = F, message = F, warning = F}
options(repos = c(CRAN = "https://cloud.r-project.org/"))
install.packages("librarian")
librarian::shelf(uuid, DBI, RMySQL, lubridate, kableExtra)
```



## Connection to the DB
```{r connect to db, eval = T, echo=T, message=F, warning=F}

# 1. Settings
db_host_db4free <- "db4free.net"
db_port_db4free <- 3306
db_name_db4free <- "pract1rrr"
db_user_db4free <- "rajivrny"
db_pwd_db4free <- "PWHWBD8RJ6"

# 2. Connect to remote server database
mydb.db4free <-  dbConnect(RMySQL::MySQL(), 
                         user = db_user_db4free, 
                         password = db_pwd_db4free,
                         dbname = db_name_db4free, 
                         host = db_host_db4free, 
                         port = db_port_db4free)
```

```{r cleardb, eval = T, echo = F, message=F, warning=F, results='hide'}
# Essentially resets all created components added to the MySQL db
dbExecute(mydb.db4free, "DROP TABLE IF EXISTS `wildlife.strikes`;")
dbExecute(mydb.db4free, "DROP TABLE IF EXISTS  flights;")
dbExecute(mydb.db4free, "DROP TABLE IF EXISTS `conditions`;")
dbExecute(mydb.db4free, "DROP TABLE IF EXISTS `wildlife_sizes`;")
dbExecute(mydb.db4free, "DROP TABLE IF EXISTS `airports`;")
dbExecute(mydb.db4free, "DROP TABLE IF EXISTS `update_log`;")
dbExecute(mydb.db4free, "DROP PROCEDURE IF EXISTS `update_wildlife_strike`;")
```

## Creating Table Schema
```{r create database schema, eval =T, echo = T, message = F, warning = F, results='hide'}

# Using VARCHAR for airports and flights to store synthetic UUIDs

# Made VARCHAR(255) to avoid BLOB/TEXT errors

# Using VARCHAR for airportCode as well because using a default value is not supported for type TEXT in db4free and all airportCodes are 3 letters so VARCHAR(3) is suitable.
dbExecute(mydb.db4free, "
    CREATE TABLE IF NOT EXISTS airports (
      aid VARCHAR(36) PRIMARY KEY,
      airportName VARCHAR(255),
      airportState VARCHAR(255),
      airportCode VARCHAR(3) DEFAULT 'ZZZ',
      UNIQUE (airportName, airportState)
     );")
    
dbExecute(mydb.db4free, "
    CREATE TABLE IF NOT EXISTS flights (
      fid VARCHAR(36) PRIMARY KEY,
      date DATE,
      originAirport VARCHAR(36),
      airlineName TEXT,
      aircraftType TEXT,
      isHeavy BOOLEAN DEFAULT FALSE,
      FOREIGN KEY (originAirport) REFERENCES airports(aid)
      );")
      
dbExecute(mydb.db4free, "
    CREATE TABLE IF NOT EXISTS `conditions` (
      cid INTEGER AUTO_INCREMENT PRIMARY KEY,
      sky_condition TEXT,
      explanation TEXT
      );")

dbExecute(mydb.db4free, "
    CREATE TABLE IF NOT EXISTS wildlife_sizes (
      wid INTEGER AUTO_INCREMENT PRIMARY KEY,
      sizes TEXT NOT NULL
    );")
      
dbExecute(mydb.db4free, "
    CREATE TABLE IF NOT EXISTS `wildlife.strikes` (
      iid INTEGER PRIMARY KEY,
      fid VARCHAR(36),
      wlsize INTEGER,
      `impact` TEXT,
      altitude INTEGER CHECK (altitude >= 0),
      `conditions` INTEGER,
      FOREIGN KEY (fid) REFERENCES flights(fid),
      FOREIGN KEY (conditions) REFERENCES conditions(cid),
      FOREIGN KEY (wlsize) REFERENCES wildlife_sizes(wid)
    );"

);
```


```{r test table existence, eval =F, echo=F, message = F, warning = F}
dbGetQuery(mydb.db4free, "SHOW COLUMNS FROM airports;")
dbGetQuery(mydb.db4free, "SHOW COLUMNS FROM flights;")
dbGetQuery(mydb.db4free, "SHOW COLUMNS FROM conditions;")
dbGetQuery(mydb.db4free, "SHOW COLUMNS FROM `wildlife.strikes`;")
dbGetQuery(mydb.db4free, "SHOW COLUMNS FROM wildlife_sizes;")
```




```{r load csv, eval = T, echo = F, message = F, warning = F}

bds.raw <- read.csv("https://s3.us-east-2.amazonaws.com/artificium.us/datasets/BirdStrikesData-V4-SuF24.csv", stringsAsFactors = FALSE)

```

## Inserting Values into Lookup Tables
```{r insert into lookup, eval =T, echo = T, message = F, warning = F, results='hide'}

# Extract unique conditions from the data frame
unique_conditions <- unique(bds.raw$sky_conditions)
insert_conditions <- paste("INSERT IGNORE INTO `conditions` (sky_condition, explanation) VALUES",
                          paste(sprintf("('%s', '')", unique_conditions), collapse = ", "),
                          ";")
dbExecute(mydb.db4free, insert_conditions)
unique_sizes <- unique(bds.raw$wildlife_size)
unique_sizes <- unique_sizes[unique_sizes != ""]
insert_sizes <- paste("INSERT IGNORE INTO `wildlife_sizes` (sizes) VALUES",
                          paste(sprintf("('%s')", unique_sizes), collapse = ", "),
                          ";")
dbExecute(mydb.db4free, insert_sizes)
```

## Populating Airports Table from Raw Data
```{r load from bds.raw to airports, eval = T, echo = T, message = F, warning = F}

# Deals with apostrophe quotes in strings within the df like "O'Hare"
apostrophe_check <- function(string) {
  gsub("'", "''", string)
}

# Defines values for string insertion
create_airport_inserts <- function(r, bds.raw) {
  uuid_value <- UUIDgenerate()
  airport_name <- apostrophe_check(bds.raw$dep_airport[r])
  state_name <- apostrophe_check(bds.raw$origin_state[r])
  paste0("('", uuid_value, "', '", airport_name, "', '", state_name, "')")
}


n.airports <- nrow(bds.raw)
batch_size <- 1000

dbBegin(conn = mydb.db4free)

#Make a loop so it does insert in batches
for(start in seq(1, n.airports, by = batch_size)) {
  end <- min (start + batch_size - 1, n.airports)
  
  # Use the named function to create the value strings
    values <- sapply(start:end, function(r) create_airport_inserts(r, bds.raw))
  
  # Write SQL statement
  sqlStmt <- paste0("INSERT IGNORE INTO airports (aid, airportName, airportState) VALUES ", 
                      paste(values, collapse = ", "))
  
  #Execute SQL Statement
  nr <- dbSendStatement(conn = mydb.db4free, statement = sqlStmt)
  dbClearResult(nr)
}

#Commits changes made since begin all at once to save time
dbCommit(conn=mydb.db4free)

```


## Populating Flights Table from Raw Data
```{r load from bds.raw to flights, eval = T, echo = T, message = F, warning = F}
bds.raw$fid <- UUIDgenerate(n = nrow(bds.raw))

convert_date_time_to_date <- function(date_time_string) {
 if (is.null(date_time_string) || is.na(date_time_string) || trimws(date_time_string) == "") {
   return(NULL)
 }
  
 date_time <- mdy_hm(date_time_string)
 return (as.Date(date_time))
}

# This finds the origin airport id by matching common fields from bds.raw
# that also exist in the airport data frame.
airports_query <- "SELECT aid, airportName, airportState FROM airports"
airports_df <- dbGetQuery(mydb.db4free, airports_query)

get_origin_airport_aid <- function(dep_airport, origin_state, airports_df) {
    matched_airport <- airports_df[apostrophe_check(airports_df$airportName) == apostrophe_check(dep_airport) & apostrophe_check(airports_df$airportState) == apostrophe_check(origin_state), ]
    return(matched_airport$aid)
}

# Defines values for string insertion
create_flight_inserts <- function(r, bds.raw) {
  fid <- bds.raw$fid[r]
  flight_date <- convert_date_time_to_date(bds.raw$flight_date[r])
  origin_airport <- get_origin_airport_aid(bds.raw$dep_airport[r], bds.raw$origin_state[r], airports_df)
  airline_name <- apostrophe_check(bds.raw$airline[r])
  aircraft_type <- bds.raw$aircraft[r]
  is_heavy <- ifelse(bds.raw$heavy_flag[r] == "Yes", TRUE, FALSE)
  
  # So I can record the null flight date values
convert_flight_date <-function(flight_date)
  if (is.null(flight_date)) {
    flight_date_sql <- "NULL" 
  } else {
    return(paste0("'", flight_date, "'")) 
  }
flight_date_sql <- convert_flight_date(flight_date)
  
  paste0("('", fid, "', ", flight_date_sql, ", '", origin_airport, "', '", airline_name, "', '", aircraft_type, "', ", is_heavy, ")")
}

n.flights <- nrow(bds.raw)
batch_size <- 5000

dbBegin(conn = mydb.db4free)

# Loop so insertions are done in batches
for(start in seq(1, n.flights, by = batch_size)) {
  end <- min(start + batch_size - 1, n.flights)
  
  values <- sapply(start:end, function(r) create_flight_inserts(r, bds.raw))
  
  #Write SQL Statement
  sqlStmt <- paste0("INSERT INTO flights (fid, date, originAirport, airlineName, aircraftType, isHeavy) VALUES ", 
                    paste(values, collapse = ", "))
  
  
  nr <- dbSendStatement(conn = mydb.db4free, statement = sqlStmt)
  dbClearResult(nr)
}
#Commit all the changes made
dbCommit(conn = mydb.db4free)
```

## Populating Wildlife Strikes Table from Raw Data
```{r load from bds.raw to `wildlife.strikes`, eval = T, echo = T, message = F, warning = F}


# So I can record the null flight date values
convert_flight_date <-function(flight_date) {
  if (is.null(flight_date)) {
    flight_date_sql <- "NULL" 
  } else {
    return(paste0("'", flight_date, "'")) 
  }
}
# Function to determine the associated cid for each sky_condition
sky_query <- "SELECT cid, sky_condition FROM `conditions`"
sky_df <- dbGetQuery(mydb.db4free, sky_query)

sky_conditions_check <- function(sky_condition, sky_df) {
    matched_sky <- sky_df[(sky_df$sky_condition) == sky_condition, ]
    return(matched_sky$cid)
}

# Function to determine the associated wid for each wildlife size
wl_query <- "SELECT wid, sizes FROM wildlife_sizes"
wl_df <- dbGetQuery(mydb.db4free, wl_query)

wl_size_check <- function(wl_size, wl_df) {
  matched_size <- wl_df[wl_size == wl_df$sizes, ]
  if (nrow(matched_size) == 0){
    return(NULL)
  }
    return(matched_size$wid)
}

# Found an issue where in rStudio the iid value when mapped to a dataframe was recorded
# as ï..iid but on posit.cloud it was recorded as iid. This is to ensure consistency.
if ("ï..iid" %in% colnames(bds.raw)) {
  colnames(bds.raw)[which(colnames(bds.raw) == "ï..iid")] <- "iid"
}

# Defines values for string insertion
create_strikes_inserts <- function(r, bds.raw) {
  iid <- bds.raw$iid[r]
  fid <- bds.raw$fid[r]
  wlsize <- wl_size_check(bds.raw$wildlife_size[r], wl_df)
  impact <- bds.raw$impact[r]
  altitude <- gsub(",","",bds.raw$altitude_ft[r])
  conditions <- sky_conditions_check(bds.raw$sky_conditions[r], sky_df)
  
  if (altitude == "" || is.na(altitude)) {
    altitude <- "NULL"
  }
  wlsize_sql <- ifelse(is.null(wlsize), "NULL", paste0("'", wlsize, "'"))
  
  
 paste0("('", iid, "', '", fid, "', ", wlsize_sql, ", '", impact, "', ", altitude, ", '", conditions, "')")
}

n.flights <- nrow(bds.raw)
batch_size <- 5000

dbBegin(conn = mydb.db4free)

# Loop so insertions are done in batches
for(start in seq(1, n.flights, by = batch_size)) {
  end <- min(start + batch_size - 1, n.flights)
  
  values <- sapply(start:end, function(r) create_strikes_inserts(r, bds.raw))
  
  #Write SQL Statement
  sqlStmt <- paste0("INSERT INTO `wildlife.strikes` (iid, fid, wlsize, impact, altitude, conditions) VALUES ", 
                    paste(values, collapse = ", "))
  
  nr <- dbSendStatement(conn = mydb.db4free, statement = sqlStmt)
  dbClearResult(nr)
}
#Commit all the changes made
dbCommit(conn = mydb.db4free)
```

## Assigning Sentinel Values for Airport Name
```{r assign sentinel values to blank airport names, eval = T, echo = T, message = F, warning = F}
 #Updating Airport name for wherever the entry is NULL or empty
dbExecute(mydb.db4free, "UPDATE airports
SET airportName = 'unknown'
WHERE airportName IS NULL OR airportName = '';")

```


## Assigning Sentinel Values for Airport State
```{r assign sentinel values to blank airport state, eval = T, echo = T, message = F, warning = F}
 #Updating Airport State for wherever the entry is N/A or empty
dbExecute(mydb.db4free, "UPDATE airports
SET airportState = 'unknown'
WHERE airportState = 'N/A' OR airportState = '';")

```


```{r show_conditions, eval = F, echo = F, message = F, warning = F}
# Show the first few rows of the conditions table
conditions_query <- dbGetQuery(mydb.db4free, "SELECT * FROM `conditions` LIMIT 5;")
print(conditions_query)
```

```{r show_wildlife_sizes, eval = F, echo = F, message = F, warning = F}
# Show the first few rows of the wildlife_sizes table
wildlife_sizes_query <- dbGetQuery(mydb.db4free, "SELECT * FROM `wildlife_sizes` LIMIT 5;")
print(wildlife_sizes_query)
```

```{r show_wildlife_strikes, eval = F, echo = F, message = F, warning = F}
# Show the first few rows of the wildlife.strikes table
wildlife_strikes_query <- dbGetQuery(mydb.db4free, "SELECT * FROM `wildlife.strikes` LIMIT 5;")
print(wildlife_strikes_query)
```

```{r show_flights, eval = F, echo = F, message = F, warning = F}
# Show the first few rows of the flights table
flights_query <- dbGetQuery(mydb.db4free, "SELECT * FROM `flights` LIMIT 5;")
print(flights_query)
```

```{r show_airports, eval = F, echo = F, message = F, warning = F}
# Show the first few rows of the airports table
airports_query <- dbGetQuery(mydb.db4free, "SELECT * FROM `airports` LIMIT 5;")
print(airports_query)
```

## Top States with Strikes

```{r top_states_with_strikes, eval = T, echo = F, message = F, warning = F}

# SQL query to find state with top 5 strike count
query <- "
  SELECT `airportState`, COUNT(*) AS `count_incidents`
  FROM `wildlife.strikes` ws
  JOIN flights f ON ws.fid = f.fid
  JOIN airports a ON f.originAirport = a.aid
  GROUP BY airportState
  ORDER BY count_incidents DESC
  LIMIT 5;"

#Get the data
top_states_with_strikes <- dbGetQuery(mydb.db4free, query)

#Display data
kable(top_states_with_strikes,
      col.names = c("State", "Number of Strikes"),
      caption = "Top 5 Steats with Wildlife Strike Incidents") %>%
  kable_styling("striped", full_width = F)
```

## Analysis by Airport

```{r analysis_by_airport, eval = T, echo = F, message = F, warning = F}
# SQL Query to find airports with below average strikes
# This selects the airports with the most incidents that are less than the average.
# The first subquery find count of incidents grouped by airport
# The second subquery finds the average incident count
# The comparison makes it so that the selected airports are only the from the first subquery
# that are less than the value from the second subquery
query_airports <- " SELECT airportName, incident_count
                    FROM (
                        SELECT a.airportName, COUNT(*) AS incident_count
                        FROM `wildlife.strikes` ws
                        JOIN flights f ON f.fid = ws.fid
                        JOIN airports a ON a.aid = f.originAirport
                        GROUP BY a.airportName
                    ) AS airport_incidents
                    WHERE incident_count < (
                      SELECT AVG(incident_count)
                      FROM(
                      SELECT COUNT(*) AS incident_count
                      FROM `wildlife.strikes` ws
                      JOIN flights f ON f.fid = ws.fid
                      JOIN airports a ON f.originAirport = a.aid
                      GROUP BY a.airportName
                    ) AS avg_incidents
                  )
                  ORDER BY incident_count DESC  
                  LIMIT 5;
                  "

#Get data
below_avg_airport <- dbGetQuery(mydb.db4free, query_airports)

#Display results
kable(below_avg_airport,
      col.names = c("Airport Name", "Number of Incidents"),
      caption = "Airports with Below Average Number of Wildlife Strike Incidents") %>%
  kable_styling("striped", full_width = F)
  
```

## Analysis by Year

```{r analysis_by_year, eval = T, echo = F, message = F, warning = F}
# Using case when to easily identify each unique wz.size and then using SUM to find the total for each size
year_query <- " SELECT YEAR(f.date) AS year, SUM(CASE WHEN wz.sizes =(SELECT DISTINCT sizes FROM wildlife_sizes LIMIT 1 OFFSET 1) THEN 1 ELSE 0 END) as small_strikes,
                                             SUM(CASE WHEN wz.sizes =(SELECT DISTINCT sizes FROM wildlife_sizes LIMIT 1) THEN 1 ELSE 0 END) as medium_strikes,
                                             SUM(CASE WHEN wz.sizes =(SELECT DISTINCT sizes FROM wildlife_sizes LIMIT 1 OFFSET 2) THEN 1 ELSE 0 END) as large_strikes
                FROM `wildlife.strikes` ws
                JOIN wildlife_sizes wz ON ws.wlsize = wz.wid
                JOIN flights f ON ws.fid = f.fid
                GROUP BY year
                ORDER BY year;"

yearly_wildlife_strikes <- dbGetQuery(mydb.db4free, year_query)

kable(yearly_wildlife_strikes,
      col.names = c("Year", "Small Strikes", "Medium Strikes", "Large Strikes"),
      caption = "Total Wildlife Strikes Grouped by Size Each Year") %>%
  kable_styling("striped", full_width = F)
```

## Trend by Year
```{r trend_by_year, eval = T, echo = F, message = F, warning = F}
xrange <- range(yearly_wildlife_strikes$year)
yrange <- range(c(yearly_wildlife_strikes$small_strikes,
                  yearly_wildlife_strikes$medium_strikes,
                  yearly_wildlife_strikes$large_strikes))


plot(xrange, yrange, type ="n", xlab = "Years", ylab="Strike Count")
colors <- c("blue", "green", "red")
linetypes <- c(1, 2, 3)
plotchars <- c(1, 2, 3)

# Defining each line
lines(yearly_wildlife_strikes$year, yearly_wildlife_strikes$small_strikes, 
      type="b", lwd=1.5, lty=linetypes[1], col=colors[1], pch=plotchars[1])
lines(yearly_wildlife_strikes$year, yearly_wildlife_strikes$medium_strikes, 
      type="b", lwd=1.5, lty=linetypes[2], col=colors[2], pch=plotchars[2])
lines(yearly_wildlife_strikes$year, yearly_wildlife_strikes$large_strikes, 
      type="b", lwd=1.5, lty=linetypes[3], col=colors[3], pch=plotchars[3])

title("Trend by Year")

# Creating a legend to easily identify lines
legend("topleft",legend=c("Small Strikes", "Medium Strikes", "Large Strikes"), 
       col=colors, pch=plotchars, lty=linetypes, cex=0.8, title="Strike Size")
```

## Creating Update Log Table and Update Procedure
```{r create_update_log, eval = T, echo = F, message = F, warning = F}
dbExecute(mydb.db4free, "CREATE TABLE IF NOT EXISTS update_log (
                            update_id INTEGER AUTO_INCREMENT PRIMARY KEY,
                            modification_type TEXT,
                            table_name TEXT,
                            time_of_update DATETIME,
                            original_values TEXT);")

# The inputs I will need to conduct the update procedure
# Inserting the update log with all the existing values of the record that will be updated
# Updating wildlife.strikes with the correct input values and using the iid passed in as an index
sp <- paste0("CREATE PROCEDURE update_wildlife_strike (
              IN strike_id INT,
              IN new_wlsize INT,
              IN new_impact TEXT,
              IN new_altitude INT,
              IN new_conditions INT)
             
             
             BEGIN
              INSERT INTO `update_log` (modification_type, table_name, time_of_update, original_values)
              SELECT 'UPDATE', 'wildlife.strikes', NOW(), CONCAT('iid: ', iid,
                                                                ', fid: ', fid,
                                                                ', wlsize: ', wlsize,
                                                                ', impact: ', impact,
                                                                ', altitude: ', altitude,
                                                                ', conditions: ', conditions)
              FROM `wildlife.strikes`
              WHERE iid = strike_id;
              
              
              UPDATE `wildlife.strikes`
              SET wlsize = new_wlsize,
                  impact = new_impact,
                  altitude = new_altitude,
                  conditions = new_conditions
              WHERE iid = strike_id;
              
              END
              "
             )
rs = dbExecute(mydb.db4free, sp)
```


## Showing Record That Will Be Updated
```{r record to be updated, eval = T, echo = F, message = F, warning = F}
# Code to check the info on the record that will be updated
to_update <- dbGetQuery(mydb.db4free, "SELECT * 
                                       FROM `wildlife.strikes`
                                       WHERE iid = 1195")
print(to_update)
```

## Calling Stored Procedure
```{r call update, eval = T, echo = F, message = F, warning = F}
# Code to Call the update
dbExecute(mydb.db4free, "CALL update_wildlife_strike(1195, 3, 'Precautionary Landing', 1000, 2)")
```


## Showing Record After Call
```{r post update, eval = T, echo = F, message = F, warning = F}
# Code to check wildlife.strikes after the update occurred to id 1195
to_update <- dbGetQuery(mydb.db4free, "SELECT * 
                                       FROM `wildlife.strikes`
                                       WHERE iid = 1195")
print(to_update)
```


## Showing Update Log After Call
```{r update recorded, eval=T, echo=F, message=F, warning =F}
# Code to check the log after the update has been entered
update_log_check <- dbGetQuery(mydb.db4free, "SELECT * 
                                       FROM `update_log`
                                       ")
print(update_log_check)
```

```{r disconnect from database, eval = T, echo = F, message = F, warning = F}
#Disconnects from the database
dbDisconnect(mydb.db4free)
```
